Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2016-07-14 pituba/article/article.tex
added new object oriented version of GITT.

% Review
%\documentclass[authoryear,preprint,review,10pt]{elsarticle}
\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}

@@

Mass balance equations for each component in fluid phase are:

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \mathbf{u}_w \cdot \mathbf{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\mathbf{u}_o \cdot \mathbf{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \mathbf{u}_o + \rho_g \mathbf{u}_g\right) \cdot \mathbf{dS}= \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.

Phase saturation is defined as its volumetric fraction of the pore space:
\begin{equation}
S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
\end{equation}
where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity, i.e., the volumetric fraction of the pore volume. Since we consider that pore volume is fully saturated with fluid,
\begin{equation}
S_o + S_g +S_w = 1 \, .
\end{equation}

A common simplification in black-oil formulations is to restrict mass transport to advection. No diffusive term is included in the mass balance equations. This simplification must be looked with care when gradients of species concentration are important. For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.

@@

%Mass balance equations for each fluid phase are \citep{ertekin2001basic}:
%% Oleo
%\begin{equation}\label{eq: So1}
%\frac{\partial}{\partial t}\left(\frac{\phi(p_P) S_o}{B_o(p_o,P_b)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_o}{B_o (p_o,P_b)}\right)=q_{os} \, ,
%\end{equation}
%% Gas
%\begin{equation}
%\begin{split}\label{eq: Sg1}
%\frac{\partial}{\partial t}\left[\phi(p_P)\left(\frac{S_g}{B_g (p_g)} +\frac{S_o R_s (p_o,P_b)}{B_o (p_o,P_b)}\right)\right] \\
%+\nabla\cdot\left(\frac{\mathbf{u}_o R_s (p_o,P_b)}{B_o (p_o,P_b)}+\frac{\mathbf{u}_g}{B_g (p_g)}\right)=q_{gs} \, ,
%\end{split}
%\end{equation}
%% Agua
%\begin{equation}\label{eq: Sw1}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_w}{B_w(p_w)}\right)=q_{ws} \, ,
%\end{equation}
%
%where saturations must satisfy the constraint
%\begin{equation}\label{eq: S11}
%S_o+S_w+S_g=1 \, .
%\end{equation}
%
%Bubble-point pressure ($P_b$) is here defined as the minimum pressure for which gas phase is locally absent, i.e., $S_g=0$. Its local value is implicitly given by the solubility ratio ($R_s$) satisfying the equation:
%\begin{equation}\label{eq: Pb}
%R_s(p_o=P_b,P_b) = \frac{B_g(p_g)}{B_o(p_o,P_b)}\frac{S_g}{S_o} +R_s(p_o,P_b) \, .
%\end{equation}
%It must be remarked that bubble-point pressure is sometimes defined in a different manner as the bubble-point pressure for the hydrocarbon composition in oil phase only. According to this latter definition, $P_b = p_o$ for any saturated reservoir.
%
%Black-oil equations make the assumption of instantaneous mass transfer between oil and gas phases, what is only reasonable for a reservoir under depletion, when oil and gas components are homonegeously distributed in oil phase. Once constant temperature and fluid composition are assumed, diffusion and advection of species and energy become unimportante for phase equilibrium calculations.
%
%If gradients of species concentration are present within oil phase, however, as it occurs in saturated reservoirs under pressure maintenance, a combined mechanism of advection and diffusion exists and those are absent of black-oil equations. Some heuristic models have been proposed for problems where this issue is critical, as for gas injection projects. \cite{utseth1981numerical} has proposed to limit $dR_s/dt$ and neglect transport of gas component in oil phase for a simplified treatment.

Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow. This extension consists in multiplying absolute (or single-phase) permeability ($\mathbf{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):

\begin{equation}\label{eq: exdarcy}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

Temperature is thus a constant prescribed value and no energy balance equation is needed. Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components. As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only. The same is valid for mass concentrations of gas and oil components in oil phase.

\begin{equation}
Y_g = Y_g (p) \, ,
\end{equation}
\begin{equation}
Y_o = Y_o (p) \, .
\end{equation}

In order to close the system of equations, a connection among different phase pressures must be provided. Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\mathbf{k},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\mathbf{k},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

%Capillary pressure and relative permeability are controversial concepts in the sense that they have not been demonstrated to arise from theoretical foundations. For practical applications, both functions are usually assumed in the forms presented above and their properties are argued to be empirically established.

%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields. Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
%
%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
%\begin{equation}
%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
%\end{equation}
%saturation discontinuities will be present due to gravity segregation.
%
%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
%\begin{equation}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \mathbf{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
%\end{equation}
%
%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\mathbf{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
%Cite Bear and Herrera.

Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume. As it will be shown, the integration process requires all quantities to be either spatially averaged or assumed constant.

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha \mathbf{g} \, .
\end{equation}
Consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
%
\subsubsection{Averaging quantities}

@@

\begin{equation}
\bar{\phi} (t) = \frac{1}{\Omega}\int_{\Omega}\phi \left(\mathbf{x},t\right) dV \, .
\end{equation}

We proceed omitting function arguments:
\begin{equation}
\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho}_w\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
\end{equation}

\item oil component:
\begin{equation}\label{eq: So3}
\frac{\partial}{\partial t} \left( \bar{\rho}_o\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}

@@

@@

This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

>>>>

% Review
%\documentclass[authoryear,preprint,review,10pt]{elsarticle}
\documentclass[final,authoryear,5p,times,twocolumn,10pt]{elsarticle}

@@

Mass balance equations for each component are:

\begin{equation}\label{eq: Sw1}
\begin{split}
\frac{\partial}{\partial t} \left( \rho_w S_w \phi \right) + \nabla\cdot\left(\rho_w \vec{u}_w \right) = \\ \rho_w^{ST}\left(Q_{wi}^{ST} - Q_w^{ST} \right) \, ;
\end{split}
\end{equation}

\begin{equation}\label{eq: So1}
\begin{split}
\frac{\partial}{\partial t}\left( \rho_o Y_o S_o \phi \right) + \nabla\cdot\left( \rho_o Y_o \vec{u}_o \right)= - \rho_o^{ST}Q_o^{ST} \, ;
\end{split}
\end{equation}

\begin{equation}\label{eq: Sg1}
\begin{split}
\frac{\partial}{\partial t}  \left[ \phi\left( \rho_o Y_g S_o + \rho_g S_g\right) \right]  + \nabla\cdot\left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right)=\\ \rho_g^{ST}\left(Q_{gi}^{ST}-Q_{g}^{ST}\right) \, .
\end{split}
\end{equation}
Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively. Source terms above account for well production or injection. Pore volume is assumed to be fully saturated, thus:
\begin{equation}
S_o + S_g +S_w = 1 \, .
\end{equation}

Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in reservoir volume ($\Omega$):

\begin{equation}\label{eq: Sw2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}\left(Q_{wi}^{ST} - Q_w^{ST} \right) - \rho_w Q_{w,aq} \, ;
\end{equation}

\begin{equation}\label{eq: So2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = - \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\begin{equation}\label{eq: Sg2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}\left(Q_{gi}^{ST}-Q_{g}^{ST}\right) \, .
\end{equation}
It was assumed that aquifer influx is the only allowed flow across reservoir boundaries.

@@

%Mass balance equations for each fluid phase are \citep{ertekin2001basic}:
%% Oleo
%\begin{equation}\label{eq: So1}
%\frac{\partial}{\partial t}\left(\frac{\phi(p_P) S_o}{B_o(p_o,P_b)}\right)+\nabla\cdot\left(\frac{\vec{u}_o}{B_o (p_o,P_b)}\right)=q_{os} \, ,
%\end{equation}
%% Gas
%\begin{equation}
%\begin{split}\label{eq: Sg1}
%\frac{\partial}{\partial t}\left[\phi(p_P)\left(\frac{S_g}{B_g (p_g)} +\frac{S_o R_s (p_o,P_b)}{B_o (p_o,P_b)}\right)\right] \\
%+\nabla\cdot\left(\frac{\vec{u}_o R_s (p_o,P_b)}{B_o (p_o,P_b)}+\frac{\vec{u}_g}{B_g (p_g)}\right)=q_{gs} \, ,
%\end{split}
%\end{equation}
%% Agua
%\begin{equation}\label{eq: Sw1}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)+\nabla\cdot\left(\frac{\vec{u}_w}{B_w(p_w)}\right)=q_{ws} \, ,
%\end{equation}
%
%where saturations must satisfy the constraint
%\begin{equation}\label{eq: S11}
%S_o+S_w+S_g=1 \, .
%\end{equation}
%
%Bubble-point pressure ($P_b$) is here defined as the minimum pressure for which gas phase is locally absent, i.e., $S_g=0$. Its local value is implicitly given by the solubility ratio ($R_s$) satisfying the equation:
%\begin{equation}\label{eq: Pb}
%R_s(p_o=P_b,P_b) = \frac{B_g(p_g)}{B_o(p_o,P_b)}\frac{S_g}{S_o} +R_s(p_o,P_b) \, .
%\end{equation}
%It must be remarked that bubble-point pressure is sometimes defined in a different manner as the bubble-point pressure for the hydrocarbon composition in oil phase only. According to this latter definition, $P_b = p_o$ for any saturated reservoir.
%
%Black-oil equations make the assumption of instantaneous mass transfer between oil and gas phases, what is only reasonable for a reservoir under depletion, when oil and gas components are homonegeously distributed in oil phase. Once constant temperature and fluid composition are assumed, diffusion and advection of species and energy become unimportante for phase equilibrium calculations.
%
%If gradients of species concentration are present within oil phase, however, as it occurs in saturated reservoirs under pressure maintenance, a combined mechanism of advection and diffusion exists and those are absent of black-oil equations. Some heuristic models have been proposed for problems where this issue is critical, as for gas injection projects. \cite{utseth1981numerical} has proposed to limit $dR_s/dt$ and neglect transport of gas component in oil phase for a simplified treatment.

\subsection{Averaging quantities}

@@

\begin{equation}
\bar{\phi} (t) = \frac{1}{\Omega}\int_{\Omega}\phi \left(\vec{x},t\right) dV \, .
\end{equation}

We proceed omitting function arguments:
\begin{equation}
\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{ \Omega}S_\alpha \phi dV \, ,
\end{equation}

@@

\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho}_w\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}\left(Q_{wi}^{ST} - Q_w^{ST} \right) - \rho_w Q_{w,aq} \, ;
\end{equation}

\begin{equation}\label{eq: So3}
\frac{\partial}{\partial t} \left( \bar{\rho}_o\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = - \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}\left(Q_{gi}^{ST}-Q_{g}^{ST}\right) \, .
\end{equation}

@@

Temperature is thus a constant prescribed value and no energy balance equation is needed. Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components. As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only. The same is valid for mass concentrations of gas and oil components in oil phase:
\begin{equation}
Y_g = Y_g (p_o) \, ,
\end{equation}
\begin{equation}
Y_o = Y_o (p_o) \, .
\end{equation}

In order to close the system of equations, a connection among different phase pressures must be provided. Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\vec{\vec{k}},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\vec{\vec{k}},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

So far, no simplification was made to black-oil equations.

Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow. This extension consists in multiplying absolute (or single-phase) permeability ($\vec{\vec{k}}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):

\begin{equation}\label{eq: exdarcy}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \vec{g}\right) \, , \quad \alpha=o,w,g \, .
\end{equation}
%Capillary pressure and relative permeability are controversial concepts in the sense that they have not been demonstrated to arise from theoretical foundations. For practical applications, both functions are usually assumed in the forms presented above and their properties are argued to be empirically established.

%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields. Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
%
%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
%\begin{equation}
%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
%\end{equation}
%saturation discontinuities will be present due to gravity segregation.
%
%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
%\begin{equation}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{\vec{k}}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
%\end{equation}
%
%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{\vec{k}}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
%Cite Bear and Herrera.

%Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume. As it will be shown, the integration process requires all quantities to be either spatially averaged or assumed constant.

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \vec{g}\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha \vec{g} \, .
\end{equation}
Consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
%

@@

This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
\begin{equation}
B_o \approx B_o^0 + \left(\frac{\partial B_o}{\partial p}\right)_T\frac{\partial p}{\partial z}dz
\end{equation}
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}
